---
title: "2026 National Square Dance Convention"
author: "Jennifer Kirk"
date: "2025-12-29"
output: html_document
---

This R Markdown file is part of Jennie Kirk’s capstone project. She contacted the 2026 National Square Dance Convention and was sent 3 CSV files with registration data for the convention to be held in Waco, TX June 24 - 27th.



#### Set up R
```{r error=FALSE}

library(tidyverse)
library(ggplot2)
library(dplyr)
#install.packages("usmap")
library(usmap)
#install.packages("ggrepel")
library(ggrepel)
#install.packages("gtools")
library(gtools)
#install.packages("UpSetR")
library(UpSetR)
```

#### Bring in CSV files

```{r}
info <- read.csv("C:/Users/jenni/Downloads/75th_Persons_Information.csv")
leaders <- read.csv("C:/Users/jenni/Downloads/75th_Dance_Leaders.csv")
pref <- read.csv("C:/Users/jenni/Downloads/75th_Dance_Preferences.csv")

# Print out the column names just to verify what they are
colnames(info)
colnames(leaders)
colnames(pref)

```



```{r}
# add a constant value to the preference dataset to use when pivoting the dataframe
pref$indic <- 1

# pivot the dataset for dance type
preft <- pref %>%
  pivot_wider(
    names_from = X.lk..Dance.Type,
    values_from = indic,
    values_fill = 0
  )
```

### Combine three dataframes into one

```{r}
combined <- info %>%
  full_join(leaders, by = "X.psn..ID") %>%
  full_join(preft, by = "X.psn..ID")

colnames(combined)

```

### Look into the data 

```{r}

# check if there are any duplicate id values
ckcombined <- combined %>% distinct(X.psn..ID)
#visually checked that there were 2342 observations in both combined and ckcombined

# check two different country variables
pivot_country <- table(combined$X.ab..Country.x,combined$X.ab..Country.y) 

print(pivot_country)

# check two different state columns  - there are too many options for the table function to be all that
# effective so subset the data where the two values do not match
pivot_state <- subset(combined,X.ab..State.Province.x !=X.ab..State.Province.y)

# add a print (there are no records)
print(pivot_state)

# take a look at the unique values
ck_state <- unique(combined$X.ab..State.Province.x)

# Note there are provinces listed and Texas appears as TX and Tx
print(ck_state)

# select columns to keep and rename them 
clean_combined0 <- combined %>% 
  select(ID = X.psn..ID,
         number_prior_conventions = X.psn..Prior.Convention,
         solo = X.psn..Solo,
         youth = X.psn..Youth,
         country = X.ab..Country.x,
         state = X.ab..State.Province.x,
         dl_caller = X.psn..Caller,
         dl_cuer = X.psn..Cuer,
         dl_lines = X.psn..Lines,
         dl_contra = X.psn..Contra,
         dl_clog = X.psn..Clog,
         refunded = X.psn..Refunded,
         type_MS = Mainstream,
         type_PLUS = Plus,
         type_CONTRA = Contra,
         type_SOLO = "Solo Dancer", 
         type_ROUND = "Round Dancing",
         type_SSD = "Social Square Dance",
         type_A = Advanced,
         type_C = Challenge,
         type_LINE = "Line Dancing",
         type_HEX = Hexagons,
         type_CLOG = Clogging,
         type_DBD = DBD,
         type_HANDI = Handicapable)

# make sure all state values that have a length of two are capitalized
clean_combined <- clean_combined0 %>%
  mutate(
    state = ifelse(str_length(state) == 2, str_to_upper(state), state)
  )
```

### Write out a cleaned combined dataset

```{r}

write.csv(clean_combined[, c("ID", "number_prior_conventions", "state", "country", "solo", "youth")], "combined.csv", row.names = FALSE)

```

### Create a dataset and map with the count and category for count by state

```{r}

# add state count
clean_combined <- clean_combined %>% 
  add_count(state, name = "state_count")

# only keep registrations from USA and one per state
statecount <- clean_combined %>% 
  filter(country == "USA") %>%
  distinct(state, state_count)

#create breaks for every 20 percent and print them
breaks <- quantile(statecount$state_count, probs = seq(0, 1, 0.20), na.rm = TRUE)
print(breaks)

# add column with values based on those breaks
statecount <- statecount %>%
  mutate(quartile_value = cut(state_count, 
                              breaks = breaks, 
                              include.lowest = TRUE, 
                              labels = c("1 - 5", "6 - 17", "18 -27", "28 - 55", "56 - 866")))

# write out the csv file
write.csv(statecount, "statecount.csv", row.names = FALSE)

# plot of the counts by state
plot_usmap(data = statecount, values = "state_count", color = "white") + 
  scale_fill_continuous(
    low = "lightblue", high = "blue", 
    name = "ID Count", label = scales::comma
  ) + 
  labs(title = "Count of IDs by State") +
  theme(legend.position = "right")

# plot the map by quartile
plot_usmap(data = statecount, values = "quartile_value", color = "white") + 
  # Use scale_fill_manual for specific custom colors
  scale_fill_manual(
    values = c("1 - 5" = "cyan1", "6 - 17" = "darkturquoise", "18 -27" = "cornflowerblue", 
               "28 - 55" = "blue3", "56 - 866" = "blue4"), 
    name = "Quartile Level",
    na.value = "snow3" # Color for states not in your data
  ) + 
  labs(title = "US States by Quartile Category") +
  theme(legend.position = "right")


```

### Take a look at the type of dances registrants perfer

```{r}

# only keep records where at least one type of dance was selected
# there were two types of registration forms so not everyone was asked
dances0 <- clean_combined |>
  select (ID, number_prior_conventions, solo, youth, country, state,
         type_MS, type_PLUS, type_CONTRA, type_SOLO, type_ROUND, type_SSD,
         type_A, type_C, type_LINE, type_HEX, type_CLOG, type_DBD, type_HANDI) |>
         filter(if_any(c(type_MS, type_PLUS, type_CONTRA, type_SOLO, type_ROUND, type_SSD,
         type_A, type_C, type_LINE, type_HEX, type_CLOG, type_DBD, type_HANDI), ~!is.na(.x)))

# pivot the type of dances preferred
dances <- dances0 %>%
  pivot_longer(
    cols = starts_with("type_"),
    names_to = "dancetype",
    names_prefix = "type_",
    values_to = "selected"
  ) %>% 
  filter(selected == 1) %>%
  mutate(dummy = "All")

# get counts by dancetype
summary_dance <- dances %>%
  count(dancetype, sort = TRUE) 

# get the total number of registrations that also responded to type of dance
totaltype <- clean_combined %>%
  filter(if_any(c(type_MS, type_PLUS, type_CONTRA, type_SOLO, type_ROUND, type_SSD,
         type_A, type_C, type_LINE, type_HEX, type_CLOG, type_DBD, type_HANDI), ~!is.na(.x))) %>% 
  count(, name = "total_participants") %>%
  mutate(dummy = "All")

# get the summary by dance type
summary_pct <- dances %>%
  group_by(dummy, dancetype) %>%
  summarize(count = n(), .groups = 'drop') %>%
  # Join the totals back in to calculate %
  left_join(totaltype, by = "dummy") %>%
  mutate(percent = (count / total_participants) * 100)

# write out csv file
write.csv(summary_pct, "dancetype.csv", row.names = FALSE)

# look at plot of preferred dance type
ggplot(data=summary_pct,aes(x = .data$dancetype, y = .data$percent)) +
  geom_col() +
  labs(title = "Preferred Dance Type",
       x="Type of Dance",
       y="Percent of Participants Selected") 

# rename the top 5 most preferred dance types
dances1 <- dances0 %>%
  rename(Mainstream = type_MS, 
         Plus = type_PLUS, 
         "Round Dancing" = type_ROUND, 
         Advanced = type_A, 
         "Line Dancing" = type_LINE)
 
# UpSetR package was built using an older version of ggplot2 logic and uses aes_string() which 
# throws some warnings so supress those warnings
suppressWarnings({
  # create an UpSet plot to show how the type of dances interact
  upset(dances1, 
        sets = c("Mainstream", "Plus", "Round Dancing", "Advanced", "Line Dancing"), 
        order.by = "freq",
        mainbar.y.label = "Number of Shared Respondents",
        sets.x.label = "Respondents per Activity",
        text.scale = c(1.5, 1.3, 1.2, 1.2, 1.5, 1.2))
})


```

#### In Jennie's experience line dancing at square dances has been informal and taught as they go so she was surprised that there were 2 registrations with only line dancing so examine that more


```{r}

# subset data for linedancing
linedance0 <- dances1[dances1$"Line Dancing" == 1,]

# go ahead and rename other dance type variables
linedance <- linedance0 %>%
  rename(Contra = type_CONTRA,
         SSD = type_SSD,
         Challenge = type_C, 
         Hex = type_HEX, 
         Clog = type_CLOG, 
         DBD = type_DBD, 
         HANDI = type_HANDI)

# there were no registrations that selected Line Dancing and either DBD nor Handi so those were left off
suppressWarnings({
  upset(linedance, 
        sets = c("Line Dancing","Mainstream", "Plus", "Round Dancing", "Advanced", "Contra", "SSD",
                 "Challenge","Hex","Clog"), 
        order.by = "freq",
        mainbar.y.label = "Number of Shared Respondents",
        sets.x.label = "Respondents per Activity",
        text.scale = c(1.5, 1.3, 1.2, 1.2, 1.1, 1.2))
})

# confirmed the two that were just line dancing also selected challenge as another dance type
```

